Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-22 latex/main.tex
Tidy up copying pattern description
Closes #698

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x;
See Fig~\ref{fig:copying_pattern_examples} for a description of these
SnpIt-like~\cite{OToole2024} plots, in which recombinants are shown
as a mosaic of variants inherited from the left vs right parents
(green vs blue, de-novo mutations in gold).}
\label{fig:averted_mutations}
\end{figure}

@@

Inspection of the copying patterns for all the recombination events
clearly identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion\cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS, of which 6 failed QC.

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
\end{tabularx}
\caption{
Illustrative copying patterns
drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}. Q1-Q4 correspond to copying patterns for recombination nodes 427863, 728991, 411345, and 663484, none of which are associated with designated Pango recombinant lineages.
The SnpIt-like~\cite{OToole2024} schematic highlights the specific genomic positions
where the allelic state in a recombinant
(middle row, labelled by node ID)
matches that of the suggested left (P0: upper row, coloured green) or right parent (P1: lower row, coloured blue),
or where the recombination event requires a de-novo mutation (gold, with mutational change at bottom).
Genome position (``pos'') is shown above each column, with the allelic state of the Wuhan-Hu-1 reference genome (``ref'') immediately below;  parental states that correspond to the reference but are not inherited by the recombinant are shown with a gray background.
Underneath each copying pattern, adjacent genomic positions are underlined in red, and
near-adjacent sites (within 3 bases) in orange.
A continuous red-and-orange underline thus indicates
a region treated as a single ``locus'' for the purposes of quality control (see Supplementary S1.XXX).
}
\label{fig:copying_pattern_examples}
\end{figure}

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, Xx 
(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

We developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci
(See Fig~\ref{fig:copying_pattern_examples} caption for details).
Inspection of these copying patterns for all the recombination events
identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion\cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS, of which 6 failed QC.

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
\end{tabularx}
\caption{
Illustrative copying patterns
drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}. 
Each copying pattern shows the 
the positions
where the allelic state in a recombinant
(middle row, labelled by node ID)
matches that of the left (P0: upper row, coloured green) 
or right parent (P1: lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
Parental states that correspond to the reference but are not 
inherited by the recombinant are shown with a gray background.
Genome position (``pos'') and reference allele (``ref'') are
shown for each column.
Underneath the copying pattern, adjacent genomic positions are 
underlined in red, and
near-adjacent sites (within 3 bases) in orange.
}
\label{fig:copying_pattern_examples}
\end{figure}
